Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-04-02 trunk/paper/response-bioinformatics.txt

Dear Dr. Bateman,

Thank you for providing us with the opportunity to revise our
manuscript. We have now prepared a revision in which we have addressed
all reviewers' comments. In response to both reviewers, we have
removed the discussion of the heart beating example from the
introduction and now provide a discussion of the axioms in the
manuscript as well as a detailed presentation of the axioms as
supplementary material. We have included a response to the individual
points raised by the reviewers below.  We would like to thank the
reviewers for their valuable comments which allowed us to improve our
manuscript.

Please do not hesitate to contact us if you require any further
information.

With best wishes,
Robert.

Responses to reviewers
----------------------

Reviewer 1:
"Introduction: I find the heart beating example difficult to follow,
primarily because it lacks simple easy-to-understand definitions - so
either this section is removed or needs to be rewritten (the result
section, however, is much clearer)."

Response: We have removed the section on heart beating from the
introduction, as recommended by the reviewer.

Reviewer 1:
p4 - On the one hand, I agree with "abnormalities in cell division
resulting in Aneuploidy refer to abnormalities of cell division
processes, while an increased rate of cell division refers to an
abnormality in the pattern of occurrence of multiple cell division
processes". However, you then state "an increased rate of cell
division is not an attribute of cell division processes, but rather an
attribute of the regulation of cell division"
two questions: 
1. is the regulation of a process(es) not different from that
collection of processes?
2. is it not the case that a collection of processes has attributes
based on its parts?

Response:
We agree with the reviewer that "regulation" processes do not always
(or commonly) include the processes that are being regulated as
parts, and we have modified our manuscript in this regard, providing
more details on regulation processes in GO and their relation to
collection processes.

Reviewer 1:
This paper needs to discuss the axiom patterns that generates the ontology.

Response:
We agree with the reviewer, and now provide information about the
axioms both in the manuscript (new section 2.3 "Axioms for physiology
phenotypes") and full details as supplementary material.

Reviewer 1:
abstract: suggests that its application is in the unification of
cellular phenotypes across species; but it's not clear where
species-specific differentiation would come into play. it also
suggests (and is further outlined in the discussion) that it could be
ussed for systems microscopy, but no evidence is provided to support
the claim. And the latter, really must be further developed, to
demonstrate that the ontology can be used for some application.

Response: 
The Cell Phenotype Ontology, which we have built by applying our
ontology-patterns for representing physiology phenotypes, is currently
being applied for the annotation of cellular observations in systems
microscopy. For this purpose, CPO provides the necessary terms (in the
sense of a controlled vocabulary) as well as a taxonomic structure
that is used for basic retrieval operations.

We agree with the reviewers that an application of our work in a
bioinformatics analysis would be highly beneficial. For example, once
sufficient data is annotated with the CPO, we can use its taxonomic
structure for enrichment analyses. Similarly, once more cellular
phenotypes are being annotated in various model organism databases, we
can utilize CPO for integrative phenotype analyses across
species. However, at the moment, no comprehensive ontology of
physiology phenotypes is available, partly due to the complexity of
the domain and partly due to limitations in available ontology
frameworks. We see our main contribution in the development of a
theory that allows to capture physiology phenotypes; we applied this
theory to construct (largely automatically) the CPO, which is now
being used to annotate cellular phenotypes in the systems microscopy
domain. Based on our contributions, it is now possible to annotate
and eventually integrate (through the use of a controlled vocabulary
of physiology phenotypes, of which the CPO provides a first example)
data in the domain of physiology. At the same time, we maintain
compatibility with GO (on which a majority of physiology phenotypes
are based) and other PATO-based phenotype ontologies, thereby filling
a gap that has limited data integration in the physiology domain so
far.

Reviewer 2:
A theory of physiology related phenotypes and an implementation based
on GO, GO cross-products and PATO is presented. An outline of a theory
of phenotypes is introduced which participate-in a set of pathological
processes which can be qualified by a small set of PATO qualities. At
least, for a majority of heterogeneous phenotype descriptions this can
be an approach for unification. In future work a sound formal
characterization has and an evaluation of the practical value of the
approach has to be delivered.

Response:
We agree, and have added a detailed formal presentation of our axioms
as supplementary material.

### 2012-04-02 trunk/paper/drinking-bioinformatics.tex

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes: the
first are process attributes that arise directly from processes and
include {\em duration} and {\em temporal location}; the second are
attributes that arise from processes and their temporal parts and
include {\em frequency} and {\em onset}; and the third are attributes
that arise from processes and qualities of their participants, and
include {\em flow rates}.

@@

However, the majority of attributes that characterize processes are
not based on these types of process attributes alone, but rather
relate attributes of process participants with the duration of a
process. In particular, a {\em rate} typically refers to an attribute
of some entity {\em with respect to an attribute of another entity},
and in the context of processes, rates often refer to attributes of a
process participant with respect to the duration of the process. For
example, a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.eps}
  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

@@

While some processes can be subdivided indefinitely while retaining
certain kinds of attributes, others cannot.
% A class $C$ of {\em continuous
%   processes} is a class that has processes $p$ as instance such that
% all temporal parts of $p$ are also instances of $C$. 
Examples of processes that can be divided include {\em continuous
  movements} or {\em mass flow} processes, for which all parts have a
{\em speed} or {\em flow rate} attribute. On the other hand, some
processes can be subdivided into stages of activity and stages of
inactivity and cannot arbitrarily be divided. For example, a process
of {\em heart beating} has periods of activity (a single heart beat)
and inactivity. Consequently, not all parts of the process have a {\em
  heart rate} attribute.

We may further attribute a {\em frequency} or {\em rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em heart beating} is
actually occurring.  This reference process can be uniquely determined
for processes such as {\em continuous movement}, where the rate of an
object at a time $t$ is the rate of the infinitesimally small process
that occurs around $t$. The reference process is ambiguous for
processes such as {\em heart beating}, and the reference process must
be explicitly stated.

@@

Amongst the abnormal phenotypes that we include for all cell
components listed in GO-CC, we distinguish {\em abnormal morphology}
and {\em abnormal physiology} phenotypes. An {\em Abnormal morphology
  of $C$} is either the (abnormal) absence of required parts of $C$,
the (abnormal) presence of additional parts, or abnormal qualities of
$C$ \cite{Hoehndorf2010phene}. For example, an absence of caveolae
({\tt MP:0004150}) would be a subclass of {\em Abnormal morphology of
  plasma membrane} in virtue of caveolae necessarily being part of the
{\em Plasma membrane} ({\tt GO:0005886}).

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of {\em Cell division}
processes, while an {\em Increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO we follow the GO and represent abnormalities in
the pattern of occurrence of $X$ as abnormalities of {\em Regulation
  of $X$} processes.  In particular, an {\em Increased rate of cell
  division} is not an attribute of {\em Cell division} processes, but
rather arises from the collection of all {\em Cell division} processes
that occur within an organism at a given time. In the
GO, {\em Regulation of $X$} processes refer to those processes that
determine how often and the way in which one or more $X$ processes
occur. Therefore, we assign the attribute of {\em Increased rate of
  cell division} to {\em Regulation of cell division} processes.

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em cytoplasmic streaming} process.

@@

The second type of abnormality in the CPO relate to abnormalities of
{\em multiple} occurrences of some process $X$. According to the GO,
{\em regulation} processes are processes that maintain or modify the
occurrence of processes of a particular type. Following this
convention, we call an abnormality of multiple occurrences of $X$ {\em
  abnormality of the regulation of $X$}.

\begin{table*}
  \centering
  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
    & Increased flow rate & Normal flow rate & Decreased
    flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Interdependency for the attribute {\em
      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
      rate} is an attribute of {\em Regulation of cytoplasmic
      streaming} processes, while {\em Cytoplasmic flow rate} is an
    attribute of individual {\em cytoplasmic streaming}
    processes. Depending both on whether the cytoplasmic flow rate in
    individual {\em cytoplasmic streaming} processes is increased or
    decreased and whether the frequency of occurrence of {\em
      cytoplasmic streaming} is increased or decreased, the total
    cytoplasmic flow rate can be increased or decreased.}
\end{table*}

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. In these
abnormalities, the {\em way} in which processes of a particular kind
are temporally distributed is abnormal.  The most common abnormal
distribution pattern is an increased or decreased frequency, and we
use PATO's {\em frequency} class to define {\em Abnormal frequency of
  occurrence of $X$}.
% \begin{verbatim}
% has-phenotype some (has-part-some (participates-in some 
%   (regulates some X and has-quality some (frequency and towards some X))))
% \end{verbatim}
For example, an {\em Abnormal frequency of occurrence of apoptosis} is
defined as an abnormality of {\em Regulation of apoptosis} ({\tt
  GO:0042981}) with respect to the {\em frequency} ({\tt
  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em onset} ({\tt PATO:0002325}) and its children
{\em delayed} and {\em premature} to define these types of regulatory
abnormality. Similarly, we use PATO's {\em offset} ({\tt
  PATO:0002324}) quality and its children to characterize regulatory
abnormalities in which a process ends prematurely or too late.

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em cytoplasmic streaming} process but
rather the total cytoplasmic flow rate, as a summation over all
cytoplasmic streaming processes that occur within an organism (or a
particular anatomical location), is increased or decreased. While a
flow rate of a single {\em cytoplasmic streaming} process is a quality
of that process, an increased {\em total} cytoplasmic flow rate is a
quality of the regulation of {\em cytoplasmic streaming}. In
particular, it is possible for an organism to have a normal --- or
even a decreased --- cytoplasmic flow rate in each individual
cytoplasmic streaming process while at the same time having an
increased total cytoplasmic flow rate due to a large increase in the
frequency of occurrence of cytoplasmic streaming processes. Similarly,
the frequency of occurrence of cytoplasmic streaming may be normal or
decreased while the total cytoplasmic flow rate is increased due to an
increased cytoplasmic flow rate in each individual cytoplasmic
streaming process. Table \ref{tbl:flow} illustrates the dependencies
between rates of individual processes, their frequency of occurrence
and the total rate of these processes. We include total rates as
regulatory abnormalities in the CPO since these are the attributes of
processes that are often measured or observed, while the rates of
individual processes are inferred following a schema such as Table
\ref{tbl:flow}.

@@

\section{Discussion}
\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

@@

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration can be achieved by further developing CPO.

This ontology will be used to integrate phenotypes' definitions across
existing datasets and will then become an integrated part of the data
processing pipeline and used to annotate the data as it gets generated
\cite{Conrad2011}.

@@

\section{Acknowledgement}
We thank Michel Dumontier for valuable discussions about the formal
representation of the axioms.

>>>>

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes: the
first are process attributes that arise directly from processes and
include {\em Duration} and {\em Temporal location}; the second are
attributes that arise from processes and their temporal parts and
include {\em Frequency} and {\em Onset}; and the third are attributes
that arise from processes and qualities of their participants, and
include {\em Flow rate}s.

@@

However, the majority of attributes that characterize processes are
not based on these types of process attributes alone, but rather
relate attributes of process participants with the duration of a
process. In particular, a {\em Rate} typically refers to an attribute
of some entity {\em with respect to an attribute of another entity},
and in the context of processes, rates often refer to attributes of a
process participant with respect to the duration of the process. For
example, a {\em Mass flow rate} refers to the {\em Mass} of a process
participant with respect to the {\em Duration} of the process, i.e.,
how much matter is moved (from one point to another) through the
process.  As a more complex example, a {\em Rate of change of
  position} refers to the {\em Distance} that an object is moved with
respect to the {\em Duration} of the process.

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em Frequency of occurrence}
or {\em Event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em Rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat rhythmically or arrhythmically within a
period of time (see Figure \ref{fig:patterns}).
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.eps}
  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em Rhythmic occurrence of P} because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em Arrhythmic occurrence of P}, and the third examples shows an
    {\em Increasing frequency of P}. A regulation process with an {\em
      Increasing frequency of P} attribute is a process in which the
    value of the {\em Frequency of occurrences of P} attribute is
    lower in the first half of the process than in the second
    half. The right side of the figure illustrates comparative
    phenotypic descriptions of processes. On the upper right, the {\em
      normal} reference process is shown. The second example
    illustrates a {\em Late onset of P}, i.e., the attribute that $P$
    processes begin later than {\em normal} processes. Finally, the
    lower right illustrates a {\em Decreased frequency of P}, since
    fewer processes of the type $P$ occur within the reference process
    than {\em normal}.}
\end{figure}

@@

While some processes can be subdivided indefinitely while retaining
certain kinds of attributes, others cannot.
% A class $C$ of {\em continuous
%   processes} is a class that has processes $p$ as instance such that
% all temporal parts of $p$ are also instances of $C$. 
Examples of processes that can be divided include {\em Continuous
  movement} or {\em Mass flow} processes, for which all parts have a
{\em Speed} or {\em Flow rate} attribute. On the other hand, some
processes can be subdivided into stages of activity and stages of
inactivity (with respect to a particular process type) and cannot
arbitrarily be divided. For example, a process of {\em Heart beating}
has periods of activity in which a heart beat occurs and periods in
which no heart beat process occurs. Consequently, not all parts of the
process have a {\em Heart rate} ({\em Rate of heart beating})
attribute.

We may further attribute a {\em Frequency} or {\em Rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart, or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em Heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em Heart beating} is
actually occurring.  This reference process can be uniquely determined
for processes such as {\em Continuous movement}, where the rate of an
object at a time $t$ is the rate of the infinitesimally small process
that occurs around $t$. The reference process is ambiguous for
processes such as {\em Heart beating}, and the reference process must
be explicitly stated.

@@

Amongst the abnormal phenotypes that we include for all cell
components listed in GO-CC, we distinguish {\em Abnormal morphology}
and {\em Abnormal physiology} phenotypes. An {\em Abnormal morphology
  of $C$} is either the (abnormal) absence of required parts of $C$,
the (abnormal) presence of additional parts, or abnormal qualities of
$C$ \cite{Hoehndorf2010phene}. For example, an {\em Absence of
  caveolae} ({\tt MP:0004150}) would be a subclass of {\em Abnormal
  morphology of plasma membrane} in virtue of caveolae necessarily
being part of the {\em Plasma membrane} ({\tt GO:0005886}).

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of {\em Cell division}
processes, while an {\em Increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO we follow the GO and represent abnormalities in
the pattern of occurrence of $X$ as abnormalities of {\em Regulation
  of $X$} processes.  In particular, an {\em Increased rate of cell
  division} is not an attribute of {\em Cell division} processes, but
rather arises from the collection of all {\em Cell division} processes
that occur within an organism at a given time. In the GO, {\em
  Regulation of $X$} processes refer to those processes that determine
how often and in which way one or more $X$ processes occur. Therefore,
we assign the attribute of {\em Increased rate of cell division} to
{\em Regulation of cell division} processes.

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em Duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em Mass flow rate}, {\em
  Energy flow rate} and {\em Velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em Cytoplasmic streaming} process.

@@

The second type of abnormality in the CPO relate to abnormalities of
{\em multiple} occurrences of some process $X$. According to the GO,
{\em regulation} processes are processes that maintain or modify the
occurrence of processes of a particular type. Following this
convention, we call an abnormality of multiple occurrences of $X$ {\em
  Abnormality of the regulation of $X$}.

% \begin{table*}
%   \centering
%   \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
%     & Increased flow rate & Normal flow rate & Decreased
%     flow rate \\
%     \hline
%     Increased frequency &increased total flow rate &increased total
%     flow rate &?\\
%     Normal frequency &increased total flow rate &normal total flow
%     rate &decreased total flow rate\\
%     Decreased frequency &?&decreased total flow rate &decreased total
%     flow rate\\
%     \hline
%   \end{tabular}
%   \caption{\label{tbl:flow}Interdependency for the attribute {\em
%       Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
%       rate} is an attribute of {\em Regulation of cytoplasmic
%       streaming} processes, while {\em Cytoplasmic flow rate} is an
%     attribute of individual {\em cytoplasmic streaming}
%     processes. Depending both on whether the cytoplasmic flow rate in
%     individual {\em cytoplasmic streaming} processes is increased or
%     decreased and whether the frequency of occurrence of {\em
%       cytoplasmic streaming} is increased or decreased, the total
%     cytoplasmic flow rate can be increased or decreased.}
% \end{table*}

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. In these
abnormalities, the {\em way} in which processes of a particular kind
are temporally distributed is abnormal.  The most common abnormal
distribution pattern is an increased or decreased frequency, and we
use PATO's {\em frequency} class to define {\em Abnormal frequency of
  occurrence of $X$}.
% \begin{verbatim}
% has-phenotype some (has-part-some (participates-in some 
%   (regulates some X and has-quality some (frequency and towards some X))))
% \end{verbatim}
For example, an {\em Abnormal frequency of occurrence of apoptosis} is
defined as an abnormality of {\em Regulation of apoptosis} ({\tt
  GO:0042981}) with respect to the {\em Frequency} ({\tt
  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em Delayed} ({\tt PATO:0000502})
or {\em Premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em Onset} ({\tt PATO:0002325}) and its children
{\em Delayed} and {\em Premature} to define these types of regulatory
abnormality. Similarly, we use PATO's {\em Offset} ({\tt
  PATO:0002324}) quality and its children to characterize regulatory
abnormalities in which a process ends prematurely or too late.

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em Cytoplasmic streaming} process but
rather the total cytoplasmic flow rate, as a summation over all
cytoplasmic streaming processes that occur within an organism (or a
particular anatomical location), is increased or decreased. While a
flow rate of a single {\em Cytoplasmic streaming} process is a quality
of that process, an increased {\em total} cytoplasmic flow rate is a
quality of the regulation of {\em Cytoplasmic streaming}. In
particular, it is possible for an organism to have a normal --- or
even a decreased --- cytoplasmic flow rate in each individual
cytoplasmic streaming process while at the same time having an
increased total cytoplasmic flow rate due to a large increase in the
frequency of occurrence of cytoplasmic streaming processes. Similarly,
the frequency of occurrence of cytoplasmic streaming may be normal or
decreased while the total cytoplasmic flow rate is increased due to an
increased cytoplasmic flow rate in each individual cytoplasmic
streaming process. 
% A {\em Total cytoplasmic flow rate} is an attribute of {\em Regulation
%   of cytoplasmic streaming} processes, while {\em Cytoplasmic flow
%   rate} is an attribute of individual {\em cytoplasmic streaming}
% processes. Depending both on whether the cytoplasmic flow rate in
% individual {\em cytoplasmic streaming} processes is increased or
% decreased and whether the frequency of occurrence of {\em cytoplasmic
%   streaming} is increased or decreased, the total cytoplasmic flow
% rate can be increased or decreased.
We include total rates as subclasses of regulatory abnormalities in
the CPO since these are the attributes of processes that are often
measured or observed, while the rates of individual processes are
inferred.

@@

\section{Discussion}
\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em Increased duration} and {\em
  Decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

@@

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration can be achieved by further developing CPO.
This ontology will be used to integrate phenotypes' definitions across
existing datasets and will then become an integrated part of the data
processing pipeline and used to annotate the data as it gets generated
\cite{Conrad2011}.

@@

\section{Acknowledgement}
We thank Michel Dumontier for valuable discussions about the formal
representation of the axioms and the different possible
interpretations of the {\bf regulates} relation.
